---
title: "Practical8"
output:
  html_document:
    df_print: paged
---

*URLs*
http://rpubs.com/adam_dennett/443357  
http://egallic.fr/R/sKDE/smooth-maps/kde.html   

*libraries*
```{r}
#install.packages("downloader")
library(tidyverse)
library(downloader)
library(rgdal)
library(sf)
library(ggplot2)
library(reshape2)
library(plotly)
#library(highcharter)
library(ggthemes)
```

Get data
```{r}
LondonWards <- readOGR("/Volumes/ucfnnap/CASA_GIS and Sc/wk8/NewLondonWard/NewLondonWard.shp", layer="NewLondonWard")
LondonWardsSF <- st_as_sf(LondonWards)
extradata <- read_csv("https://www.dropbox.com/s/qay9q1jwpffxcqj/LondonAdditionalDataFixed.csv?raw=1")
```

Merge datasets
```{r}
LondonWardsSF <- merge(LondonWardsSF, extradata, by.x = "WD11CD", by.y = "Wardcode")
```

```{r}
summary(LondonWardsSF)
names(LondonWardsSF)
```

London data store for data key=value reference
https://londondatastore-upload.s3.amazonaws.com/instant-atlas/borough-profiles/atlas.html  

Simple histogram
```{r}
### style
th <- theme_tufte(base_family = "Georgia")

### AvgPubTrans = Average Public Transport Accesibility score, 2014
ggplot(LondonWardsSF, aes(AvgPubTran)) +
  geom_histogram() + th
```

Simple histogram binwidth = 0.1
```{r}
ggplot(LondonWardsSF, aes(AvgPubTran)) +
  geom_histogram(binwidth = 0.1) + th
```

Histogram with vertical lines Central Tendency
```{r}
CT <- ggplot(LondonWardsSF, aes(AvgPubTran)) +
  geom_histogram(binwidth = 0.1) + th +
  geom_vline(aes(xintercept = mean(AvgPubTran)), colour="yellow") +
  geom_vline(aes(xintercept = median(AvgPubTran)), colour="magenta") 
```

Calulate Mode
```{r}
### Rounded PTAL to get 'most typical' value (otherwise all are unique values)
x <- round(LondonWardsSF$AvgPubTran, digits = 1)
y <- table(x)
names(y)[which(y==max(y))]
```

Histogram with Mode
```{r}
CT + geom_vline(aes(xintercept = 2.4), colour="cyan")
```

Annotate Histogram
```{r}
m_ean <- annotate("text", label = paste("Mean\n", round(mean(LondonWardsSF$AvgPubTran), digits = 2)), x = mean(LondonWardsSF$AvgPubTran), y = 7, color = "white", size = 3, family = "Georgia", hjust =-.01)
m_edian <- annotate("text", label = paste("Median\n", round(median(LondonWardsSF$AvgPubTran), digits = 2)), x = median(LondonWardsSF$AvgPubTran), y = 3, color = "white", size = 3, family = "Georgia", hjust =-.01)
m_ode <- annotate("text", label = paste("Mode: ", 2.4), x = 2.4, y = 3, color = "white", size = 3, family = "Georgia", angle = 90, vjust =-.5)

CT + geom_vline(aes(xintercept = 2.4), colour="cyan") + m_ean + m_edian + m_ode
```

Kernel density smoothed frequency gradients
```{r}
ggplot(LondonWardsSF, aes(x=AvgPubTran, y=..density..)) + geom_histogram(aes(y = ..density..))+
  geom_density(colour="red", fill="red", alpha=0.2) + 
  geom_density(colour="orange", fill="orange", alpha=0.2, adjust = 1/5) + 
  geom_density(colour="purple", fill="purple", alpha=0.2, adjust = 1/3) + th
```

Different Kernels
```{r}
ggplot(LondonWardsSF, aes(x=AvgPubTran, y=..density..)) + geom_histogram(aes(y = ..density..))+
  geom_density(colour="red", fill="red", alpha=0.2, adjust = 1/3, kernel="gaussian") + 
  geom_density(colour="orange", fill="orange", alpha=0.2, adjust = 1/3, kernel="rectangular") + 
  geom_density(colour="purple", fill="purple", alpha=0.2, adjust = 1/3, kernel="cosine") + th
```

Boxplot
```{r}
bp <- ggplot(LondonWardsSF, aes(1, AvgPubTran)) + theme_tufte() +
  geom_tufteboxplot()
bp
```

Annotate Boxplot
```{r}
fn <- fivenum(LondonWardsSF$AvgPubTran)
an <- annotate(geom="text", label=round(fn ,digits=2), x= 1, y=fn, size = 3, family = "Georgia", hjust =-.2)
bp + an + theme(axis.title=element_blank(),axis.ticks.x = element_blank(), axis.text.x=element_blank())
```

Faceted grid
```{r}
#check which variables are numeric first
list1 <- as.data.frame(cbind(lapply(LondonWardsSF, class)))
list1 <- cbind(list1, seq.int(nrow(list1)))

#you will notice that there are some non-numeric columns, we want to exclue these, and drop the geometry 
LondonSub <- LondonWardsSF[,c(1:73,83:86)]
#make sure the geometry is null or we will get errors - also create some subsets so that we can see our data better
LondonSub2 <- st_set_geometry(LondonWardsSF[,c(1:3,9:27)],NULL)
LondonSub3 <- st_set_geometry(LondonWardsSF[,c(1:3,28:50)],NULL)
LondonSub4 <- st_set_geometry(LondonWardsSF[,c(1:3,51:73,85:86)],NULL)

LondonMelt2 <- melt(LondonSub2, id.vars = 1:3)
attach(LondonMelt2)
hist2 <- ggplot(LondonMelt2, aes(x=value)) + geom_histogram(aes(y = ..density..)) + geom_density(colour="cyan", size=1, adjust=1) + theme_tufte()
hist2 + facet_wrap(~ variable, scales="free")
```

```{r}
LondonMelt3 <- melt(LondonSub3, id.vars = 1:3)
attach(LondonMelt3)
hist3 <- ggplot(LondonMelt3, aes(x=value)) + geom_histogram(aes(y = ..density..)) + geom_density(colour="cyan", size=1, adjust=1) + theme_tufte()
hist3 + facet_wrap(~ variable, scales="free")
```

```{r}
LondonMelt4 <- melt(LondonSub4, id.vars = 1:3)
attach(LondonMelt4)
hist4 <- ggplot(LondonMelt4, aes(x=value)) + geom_histogram(aes(y = ..density..)) + geom_density(colour="cyan", size=.7, adjust=1) + theme_tufte()
hist4 + facet_wrap(~ variable, scales="free")
```

Log10() transformation on the x variables
```{r}
hist5 <- ggplot(LondonMelt4, aes(x=log10(value))) + geom_histogram(aes(y = ..density..)) + stat_function(fun=dnorm, colour="cyan", size=0.5) + theme_tufte()
hist5 + facet_wrap(~ variable, scales="free")
```

spatial Kernel Density Estimate
```{r}
londonpoint<-ggplot(LondonSub, aes(x=x.y,y=y.y))+geom_point()+coord_equal()+theme_tufte()
londonpoint
```

```{r}
library(viridis)
londonpoint<-ggplot(LondonSub, aes(x=x.y,y=y.y))+stat_bin2d(bins=10)+theme_tufte()+scale_fill_viridis()
londonpoint
```

```{r}
londonpoint<-ggplot(LondonSub, aes(x=x.y,y=y.y))+geom_point()+coord_equal()+theme_tufte()
londonpoint
```

```{r}
londonpoint+stat_density2d(aes(fill = ..level..), geom="polygon")+theme_tufte()+scale_fill_viridis()
```

Introduction to functions in R  
Structure of a fucntion  
```{r}
myfunction <- function(arg1, arg2, ... ){
  statements
  return(object)
}
```

Re-classiy data (recode data)  
```{r}
newvar<-0
recode<-function(variable,high,medium,low){
  newvar[variable<=high]<-"High"
  newvar[variable<=medium]<-"Medium"
  newvar[variable<=low]<-"Low"
  return(newvar)
}

# First we initialise a new variable called newvar and set it to = 0. We then define a new function called recode. This takes in 4 pieces of information: A variable (called variable but I could have called it anything) and three values called high, medium and low. It outputs a value to the new string variable newvar based on the values of high, medium and low that are given to the function.
```

We can now use this function to recode any of our continuous variables into high, medium and low values based on the values we enter into the function.

```{r}
summary(LondonWardsSF$AvgGCSE201)
```

```{r}
LondonWardsSF$GCSE_recode <- recode(LondonWardsSF$AvgGCSE201,409.1,358.3,332.3)
```

Location Quotient
```{r}
#Location Quotient function 1
LQ1<-function(pctVariable){
  pctVariable / mean(pctVariable)
}

#Location Quotient function 2
LQ2<-function(variable,rowtotal){
  localprop<-variable/rowtotal
  globalprop<-sum(variable)/sum(rowtotal)
  return(localprop/globalprop)
}
```

Calculate Location Quotients for the 5 Housing tenure variables (Owner Occupied, Private Rent, Social Rent, Shared Ownership, Rent Free)
```{r}
names(LondonWardsSF)
```


```{r}
LondonWardsSF$Owner_occ <- LQ1(LondonWardsSF$PctOwned20)
LondonWardsSF$PRent <- LQ1(LondonWardsSF$PctPrivate)
LondonWardsSF$SRent <- LQ1(LondonWardsSF$PctSocialR)
LondonWardsSF$PShared <- LQ1(LondonWardsSF$PctSharedO)
LondonWardsSF$RFree <- LQ1(LondonWardsSF$PctRentFre)
```

Location Quotient Mapper function by A.Dennett
```{r}
#############################################################
##A Function for creating various location quotient maps
##
##By Adam Dennett October 2014 - updated November 2018
##
##Please note, this function requires input data to already be in ##the form of row percentages. To create the function, highlight the ##whole block of code and run it. To run the function, simply use  ##LQMapper(your_dataframe)

library(rgeos)
library(ggplot2)
library(maptools)
library(sf)
library(tmap)

sfdataframe <- LondonWardsSF

LQMapper<-function(sfdataframe){
  print(colnames(sfdataframe))  
  vars<-readline("From the list above, select the variables 
                 you want to calculate location quotients for 
                 separated by spaces...")
  
  # split the string at the spaces  
  vars<-unlist(strsplit(vars, split = "\\s"))  
  # now save vars as a list
  vars<-as.list(vars)  
  
  print("looping to create new location quotient variables...")
  attach(sfdataframe)  
  for(i in 1:length(vars)){
    pctVariable<-vars[[i]]
    colvect<-which(colnames(sfdataframe)==vars[[i]])
    
    #this is a little function to calculate location quotients
    LQ<-function(pctVariable){
      pctVariable/mean(pctVariable)
    }
    #use LQ function here to create new variable in sfdataframe 
    #and save it
    v <- sfdataframe[,colvect]
    sfdataframe[,paste("LQ_",pctVariable, sep="")] <- LQ(v[[pctVariable]])    
  }
  
  #reset i as we're going to use it again in a minute
  i=0
  
  print("now entering the plotting loop")
  for(i in 1:length(vars)){
    print("I'm plotting")
    pctVariable<-paste("LQ_",vars[[i]],sep="")
    colvect<-which(colnames(sfdataframe)==paste("LQ_",vars[[i]],sep=""))
    
    #create the plot
    LQMapperPlot <- tm_shape(sfdataframe) + tm_polygons(pctVariable, 
                        style="jenks",
                        palette="Spectral",
                        midpoint=1,
                        title=pctVariable,
                        alpha = 0.5)
    
    LQMapperPlot
    #save the plot to a pdf and give it a name based on its variable
    tmap_save(LQMapperPlot, filename=paste(pctVariable,".png",sep=""))
    
  }  
  return(sfdataframe)  
}

###################################################################
```

```{r}
#LQMapper(LondonWardsSF)
```

```{r}
tm_shape(LondonWardsSF) + 
  tm_polygons("SRent", 
              style="jenks",
              palette="PuRd",
              midpoint=1, 
              title="PctSocialR", 
              border.col="white",
              border.alpha = 0.01)
```

Basic Geodemographic Classification  
In a cluster analysis, you should select variables that are:
* Ranged on the same scale
* Normally distributed
* Not highly correlated

```{r}
LondonWardsDF <- st_set_geometry(LondonWardsSF, NULL)
#display list nad class of variables 
cbind(lapply(LondonWardsDF, class))
```

Create new Dataframe with 2 variables 
```{r}
GD<-as.data.frame(LondonWardsDF[,c("CarsPerHH2","PctAged65p")])
```

```{r}
mean(LondonWardsDF$CarsPerHH2) 
median(LondonWardsDF$CarsPerHH2) 
```
```{r}
mean(LondonWardsDF$PctAged65p) 
median(LondonWardsDF$PctAged65p) 
```

```{r}
# run a k-means to find 3 clusters – use 25 iterations
fit <- kmeans(GD, 3, nstart=25) # 3 cluster solution
# get cluster means
centroid<-aggregate(GD,by=list(fit$cluster),FUN=mean)
#print the results of the cluster groupings
centroid
```

```{r}
# as we only have variable two dimensions we can plot the clusters on a graph
p <- ggplot(GD,aes(CarsPerHH2, PctAged65p))
p+geom_point(aes(colour=factor(fit$cluster)))+geom_point(data=centroid[,2:3],aes(CarsPerHH2, PctAged65p), size=7, shape=18)+ theme(legend.position="none")
```

```{r}
GD$cluster <- fit$cluster

#add the cluster groups to the LondonWards data frame
LondonWardsSF$cluster<-GD$cluster
#now map our geodeomographic classification
map <- ggplot(LondonWardsSF) + geom_sf(mapping = aes(fill=cluster))+theme_tufte()+scale_fill_viridis()
map
```

